Clustering of Distributions on 1-Dimensional Manifolds

### 2021-10-13 scratch/distributions-on-1d-manifolds.Rmd
updated document

@@

### General case

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 4, 
                      fig.width = 4)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

@@

### Test for intersecting manifolds

\begin{theorem}
Let $\mathcal{M}_1$ and $\mathcal{M}_2$ be two one-dimensional manifolds of unit length with $n / 2$ points drawn uniformly along each manifold.

Suppose we want to perform a hypothesis test for whether the ends of the manifolds meet. 
The $H_0$ is the case where $\delta = 0$ and $H_1$ is the case where $\delta > 0$. 
We define the test statistic $T_n$ as follows: 
Construct an $\epsilon$-neighborhood graph using the smallest $\epsilon$ required to make the graph connected. 
Let $\{d_i\}$ be the set of edge lengths of this graph. 
Define $T_n = \max_i d_i$. 
Then $T_n / 2 \stackrel{d}{\to} \Betadist(1, n)$ under $H_0$. 
\end{theorem}

This theorem isn't 100\% correct---in reality, we want to draw edges along the "super-manifold" that is created by joining the two manifolds (under $H_0$). But if $H_0$ is true, then these lengths should converge to the true distances along the super-manifold.

### Community detection for "non-affine" one-dimensional manifolds

For non-intersecting manifolds, distance-based graph techniques provide a straightforward method for community detection. 
However, this is not the case in many models with community structure. 
For example, the latent structure that induces the degree-corrected block model consists of line segments that emanate from the origin, so each community is a one-dimensional subspace, or linear manifold, and these manifolds meet at the origin. 
For the DCBM, a straightforward approach to community detection involves an ASE followed by clustering based on angles (e.g., cosine similarity). 
As an extension to this case, we consider community detection for two one-dimensional manifolds that meet at the origin.

The method that is proposed is as follows:

1. For each point $x_i$, compute $\|x_i\|$.
2. Construct an $\epsilon$-neighborhood graph for $X$, choosing $\epsilon$ large enough that the graph is connected. 
Alternatively, construct a $K$-NN graph, again, choosing $K$ so that the graph is connected. 
3. Remove the vertex $v_j$ that corresponds to the point $x_j = \min_i \|x_i\|$. 
If this results in two disconnected subgraphs, we are done. 
Otherwise, continue removing vertices in ascending order of magnitude until the graph is disconnected. 
Then map each component to a community.
4. Estimate curves for each community (e.g., using polynomial regression). 
Map each removed point to a community based on its distance to the curve.

In the following example, we draw $64$ points each uniformly along two manifolds in $\mathbb{R}^2$ that intersect at the origin:

```{r}
import::from(magrittr, `%>%`)

source('~/dev/pabm-grdpg/functions.R')
source('https://mtrosset.pages.iu.edu/Courses/675/graph.r')

f1 <- function(t) {
  x1 <- t / sqrt(.5 + sqrt(5) / 2)
  x2 <- x1 ** 2
  return(cbind(x1, x2))
}

f2 <- function(t) {
  x1 <- t / (.5 + sqrt(5) / 2)
  x2 <- x1 ** .5
  return(cbind(x1, x2))
}

set.seed(1)
```

```{r}
n1 <- 2 ** 6
n2 <- n1
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))

a <- 1
b <- 1
t1 <- rbeta(n1, a, b)
t2 <- rbeta(n2, a, b)

X1 <- f1(t1)
X2 <- f2(t2)
X <- rbind(X1, X2)
plot(X, col = z, xlim = c(0, 1), ylim = c(0, 1))

r <- apply(X, 1, function(x) sum(x ** 2))
order.r <- order(r)
```

The $5$-NN graph is as follows:

```{r}
W <- X %>% 
  dist() %>% 
  as.matrix() %>% 
  graph.knn(5)
W <- W + t(W)
W <- sign(W)
diag(W) <- 0

qgraph::qgraph(W, groups = as.factor(z))
```

Removing the 10 points closest to the origin results in the following:

```{r}
remove <- which(order.r <= 10)
```

```{r}
W <- W[-remove, -remove]
qgraph::qgraph(W, groups = as.factor(z[-remove]))
```

## General case

### 2021-10-13 scratch/distributions-on-1d-manifolds.tex
updated document

\begin{theorem}
Let $\mathcal{M}_1$ and $\mathcal{M}_2$ be two one-dimensional manifolds of unit length with $n_1$ and $n_2$ points drawn uniformly along them ($n = n_1 + n_2$).

Suppose we want to perform a hypothesis test for whether the ends of the manifolds meet. 
The $H_0$ is the case where $\delta = 0$ and $H_1$ is the case where $\delta > 0$. 
Define the test statistic $T_n = \|X_{(n_1)} - Y_{(1)}\|$ (or similar, depending on which ends meet). 
Then $T_n / 2 \stackrel{d}{\to} \mathrm{Beta}(1, n)$ under $H_0$. 
\end{theorem}

@@

\hypertarget{community-detection-for-manifolds-that-emanate-from-the-origin}{%
\subsubsection{Community detection for manifolds that emanate from the
origin}\label{community-detection-for-manifolds-that-emanate-from-the-origin}}

>>>>

\begin{theorem}
Let $\mathcal{M}_1$ and $\mathcal{M}_2$ be two one-dimensional manifolds of unit length with $n / 2$ points drawn uniformly along each manifold.

Suppose we want to perform a hypothesis test for whether the ends of the manifolds meet. 
The $H_0$ is the case where $\delta = 0$ and $H_1$ is the case where $\delta > 0$. 
We define the test statistic $T_n$ as follows: 
Construct an $\epsilon$-neighborhood graph using the smallest $\epsilon$ required to make the graph connected. 
Let $\{d_i\}$ be the set of edge lengths of this graph. 
Define $T_n = \max_i d_i$. 
Then $T_n / 2 \stackrel{d}{\to} \mathrm{Beta}(1, n)$ under $H_0$. 
\end{theorem}

This theorem isn't 100\% correct---in reality, we want to draw edges
along the ``super-manifold'' that is created by joining the two
manifolds (under \(H_0\)). But if \(H_0\) is true, then these lengths
should converge to the true distances along the super-manifold.

\hypertarget{community-detection-for-non-affine-one-dimensional-manifolds}{%
\subsubsection{Community detection for ``non-affine'' one-dimensional
manifolds}\label{community-detection-for-non-affine-one-dimensional-manifolds}}

For non-intersecting manifolds, distance-based graph techniques provide
a straightforward method for community detection. However, this is not
the case in many models with community structure. For example, the
latent structure that induces the degree-corrected block model consists
of line segments that emanate from the origin, so each community is a
one-dimensional subspace, or linear manifold, and these manifolds meet
at the origin. For the DCBM, a straightforward approach to community
detection involves an ASE followed by clustering based on angles (e.g.,
cosine similarity). As an extension to this case, we consider community
detection for two one-dimensional manifolds that meet at the origin.

The method that is proposed is as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  For each point \(x_i\), compute \(\|x_i\|\).
\item
  Construct an \(\epsilon\)-neighborhood graph for \(X\), choosing
  \(\epsilon\) large enough that the graph is connected. Alternatively,
  construct a \(K\)-NN graph, again, choosing \(K\) so that the graph is
  connected.
\item
  Remove the vertex \(v_j\) that corresponds to the point
  \(x_j = \min_i \|x_i\|\). If this results in two disconnected
  subgraphs, we are done. Otherwise, continue removing vertices in
  ascending order of magnitude until the graph is disconnected. Then map
  each component to a community.
\item
  Estimate curves for each community (e.g., using polynomial
  regression). Map each removed point to a community based on its
  distance to the curve.
\end{enumerate}

In the following example, we draw \(64\) points each uniformly along two
manifolds in \(\mathbb{R}^2\) that intersect at the origin:

\begin{center}\includegraphics{distributions-on-1d-manifolds_files/figure-latex/unnamed-chunk-2-1} \end{center}

The \(5\)-NN graph is as follows:

\begin{center}\includegraphics{distributions-on-1d-manifolds_files/figure-latex/unnamed-chunk-3-1} \end{center}

Removing the 10 points closest to the origin results in the following:

\begin{center}\includegraphics{distributions-on-1d-manifolds_files/figure-latex/unnamed-chunk-5-1} \end{center}

@@
